---
title: "Workshop 1: Working with R"
author: "Lauren Kennedy"
date: "09/05/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Key learning outcomes

- Download the example dataset that we will use for the next two days

- Work with projects and rmarkdown documents

- Explore data munging with the tidyverse

Stretch goal: Create visualizations using example code and ggplot2

# Step one: Setting up for reproducibility

- Create a new project in R 

- Within that project, create folders for data and code

- Create a new rmarkdown document and save it in the code folder


# Step two accessing our example data 

We will be using raw data from an online convenience sample completing psychology tests online. Access data here https://openpsychometrics.org/_rawdata/ and select this data from the sixth cell down "Big Five Personality Test". 

This data is used as an example in this paper https://arxiv.org/abs/1906.11323. We will be using it as example data to explore survey weighting and multilevel regression and poststratification. 

This is an **example dataset**. It's useful for educational purposes because it is freely available and has a number of features that are useful to discuss. This tutorial is NOT intended as a claim that all web samples can be generalized by poststratification with two variables. n

Download this data, unzip the .csv save in the data folder. 

# Step three 

In the first chunk of your rmarkdown document, load the tidyverse library, the ggplot2 library and the readr library. 

Then in the second chunk read in your data. If you have set up the project as described, the following code should be sufficient. Note that instead of using read.csv, I have instead used read_table2. This is because although the data is saved as a .csv (comma separated), it is in fact white space seperated. Using the gui viewer it was easy to diagnose and solve this issue. 

```{r, eval = FALSE}
library(readr)

ex_data <- read_table2("../data/data.csv")

```

# Step four: Data exploration

Now we can R to explore the data to understand what we are working with. Here are some helpful functions:

```{r, eval = FALSE}
summary(ex_data)

View(ex_data)

head(ex_data)

tail(ex_data)
```

One important thing to check is the type of variable R thinks is in each column. You can do this with `str`. 

```{r, eval = FALSE}
str(ex_data)
```

One thing to notice is that the demographics are recorded as either numeric (num) or character ("chr"). For our later work, we would like them to be a special type of variable in R called factor. We can change these variables like this:

```{r, eval = FALSE}
library(tidyverse)

ex_data <- ex_data %>%
  mutate_at(c("race","engnat","gender","hand","source","country"), as.factor)

```

Try running summary of the data again to see how changing the variable type changes how the function works.

Another option would be to mutate each variable seperately so you can name the levels from the codebook using the `levels` arguement in the as.factor function. 

# Step five: Create new variables

We also need to create derived variables. We will need to create an age_group variable, but we'll leave this for next tutorial. Right now we will focus on making an aggregate score for extroversion - which will be a linear combination of the ten items that start with E. Extroversion = E1 + (5-E2) + E3 + (5-E4) + E5 + (5-E6) + E7 + (5-E8) + E9 + (5-E10). Items 2,4,6,8 and 10 are reverse coded, which is why we need to subtract them from 5. 

# Step six: Getting comfortable with data manipulation 

We'll have to do some more manipulation with data in later sections. Here are some exercises to practice. 

- Create a dataset of only US respondents. 

- Create a dataset of respondents who are aged 18 and older

- What is the average level of extroversion for different types of handidness

- What proportion of respondents came from a url with .edu?

- Can you pivot the data so that each item response is represented by one row?

# Step seven: A few easy visualization examples with ggplot2

Here are a few visulisations to get you started exploring the data. Feel free to ask for more if you have specific ones in mind!

```{r, eval = FALSE}
library(tidyverse)

# scatter plot
ex_data %>%
  filter(age<100 & age>0& A1>0)%>% # removing a few coding errors
  ggplot(.,aes(x = E1, y =age)) +
  geom_jitter() #like a scatter plot but with points jittered to reduce overplotting

# add colour

# scatter plot
ex_data %>%
  filter(age<100 & age>0& A1>0)%>% # removing a few coding errors
  ggplot(.,aes(x = E1, y =age, colour = as.factor(hand))) +
  geom_jitter(alpha = .3) #like a scatter plot but with points jittered to reduce overplotting

# histogram

ex_data %>%
  ggplot(.,aes(x = E1)) +
  geom_histogram()

# histogram across handidness

ex_data %>%
  ggplot(.,aes(x = E1)) +
  geom_histogram() +
  facet_grid(~hand)


```

